---
title: "How crisis affect the low-income and high-income economies differently"
subtitle: "An example from the 2020/2021 Covid-19"
author: 
  - Sean Liu 
thanks: "Code and data are available at: https://github.com/Sluuu/crisis-affect-countries.git"
date: today
date-format: long
abstract: "This paper examines the impacts of global crises focusing on the 2020 and 2021 COVID-19 pandemic's effects on low and high-income countries. It highlights how globalization links economies, leading to shared challenges during the worldwide pandemic. The analysis shows that low-income countries have suffered more than high-income countries, by experiencing higher unemployment, inflation, and slower GDP growth."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(readr)
library(dplyr)
library(arrow)
library(dataverse)
library(WDI)
library(rstanarm)
library(broom)
library(stargazer)
```


# Introduction

Globalization improves the connection of different countries of the world. Economically,  globalization benefits different countries in multiple ways and being able to trade goods across borders.[@globalization] However, the effects of globalization brings the whole economy on the same boat. Consequently, when various global crises occur, they have multiple and different impacts on the world economy.

In this paper, my focus is on the different impacts of global crises on low and high income countries. I have selected the 2020 and 2021 COVID-19 pandemic as the primary subject of my study. This crises have led to varying degrees of economic uncertainty and fluctuations in economic performance worldwide.[@compare] The World Bank indicates that the COVID-19 pandemic has widen the inequality between low and high-income countries, with low-income countries experiencing higher unemployment and inflation rates, also with slower GDP growth. These trends align with the expectations set in my study.[@inequality]
```{r}
#| echo: false
#| warning: false
#| message: false

#load data
data_na<- read_csv(here::here("data/analysis_data/analysis_data.csv"))
#dd_test <- read_parquet(here::here("data/analysis_data/data_na.parquet"))
```

## Defining Low-income economies and High-income economies

According to The World Bank [@WDIincome], as of July 2019, using the Gross National Income, GNI, per capita by using the World Bank Atlas method. The Atlas method smooths exchange rate fluctuations using a three-year moving average. The low-income economies are defined as \$1,025 or less, lower middle-income economies are those with between \$1,026 and \$3,995. The upper middle-income economies are those between \$3,996 and \$12,375. Lastly, the high-income economies are those with a GNI per capita \$12,376 or more.

For simplicity, I will group the low-income economies and the lower middle-income economies together as low-income economies. Also, the upper middle-income economies and high-income economies will be grouped as high-income economies. Therefore, a country that has a GNI per capita less than or equal to \$3995 will be categorized as low-income country. A country that has a GNI per capita larger than \$3995 will be categorized as high-income country. Which I have grouped out in 2020 there are 27 low-income countries and 85 high-income countries in this definition. In 2021, there are 26 low-income countries and 85 high-income countries. 

\newpage

# Data

The data is obtained from the World Development Indicators and Other World Bank Data, WDI,  @WDI. The raw data set includes, 17024 observations and 11 variables. As the raw data-set recorded some variables have missing values and some repeated values. After combining and deleting the extra and missing values, In the year of 2020, I have 112 observations and in the year of 2021 I have 117 variables which both years with the sharing 10 variables. Which each observations represents a country. 

## Variables 
```{r}
#| echo: false
#| warning: false
#| message: false

#Adding a new variable
data_na <- data_na %>%
  mutate(low_income = as.numeric(gni_per_cap <= 3995))
```

- `country`: It records the country's name.
- `year`: It records the year of the data being recorded. In our cleaned data set there will only be year 2020.
- `inflation`: It records the inflation, measured by the consumer price index which reflects the annual percentage change in the cost to the average consumer of acquiring a basket of goods and services that may be fixed or changed at specified intervals, such as yearly.
- `gdp_growth`: It records the annual percentage growth rate of GDP at market prices. The calculations are based on constant local currency.
- `population`: It records the total population based on all residents regardless of legal status or citizenship.
- `unem_rate`: It records the unemployment rate, the unemployment refers to the labours that are not working but is available to work or seeking for employment. 
- `lf_par_rate`: It records the labor force participation rate, it calculated the proportion of the population who ages 15 and older that are active economically. 
- `trade_per`: It records the sum of exports and imports of goods and services measured as a share of GDP.
- `gni_per_cap`: It records the GNI per capita, the unit is converted to U.S. dollars using the World Bank Atlas method. GNI is the sum of value added by all resident producers plus any product taxes. It does not include the value of output plus net receipts of primary income from abroad.
- `low_income`: It is a dummy variable indicating that if a country has a GNI per capita less than or equal to \$3995 it will be 1, else, 0.

```{r}
#| label: fig-data
#| fig-cap: Relationship between GDP growth versus Inflation
#| echo: false
#| warning: false
#| message: false

data_na |>
  ggplot(mapping = aes(x = gdp_growth, y = inflation, color = as.factor(year))) +
  geom_point() +
  theme_minimal() +
  labs(x = "GDP Growth", y = "Inflation", color = "Years")
```

In @fig-data, I have presented the relationship between GDP growth and Inflation of each countries from my data-set. Which each dot represents one country. From @fig-data, we could see that countries that have higher/positive gdp growth typically also has a higher inflation rate. Also in 2021, the gdp growth for each country are typically higher than in 2020 with having similar inflation rate. 

\newpage

```{r}
#| label: fig-inflation
#| fig-cap: Relationship between differenct level of income countries versus Inflation
#| echo: false
#| warning: false
#| message: false

data_na |>
  ggplot(mapping = aes(x = low_income, y = inflation, color = as.factor(year))) +
  geom_point() +
  theme_minimal() +
  labs(x = "Low-income countries", y = "Inflation", color = "Years")+
  scale_x_continuous(breaks = c(0,1))
```
In @fig-inflation, I have shown the inflation rate for the low-income and the high-income countries in 2020 and 2021. On the x-axis, when x is equal to 0, it means that the county is a high-income country and it equals to 1 when the country is a low-income country. From @fig-inflation, we could also see that high income countries has lower and more stabilized inflation rates in either year 2020 and year 2021. Which in low-income countries the inflation rate varies across each country and there are also more outliers in year 2021 compare with 2020. 

\newpage

```{r}
#| label: fig-gdp-growth
#| fig-cap: Relationship between differenct level of income countries versus gdp growth
#| echo: false
#| warning: false
#| message: false

data_na |>
  ggplot(mapping = aes(x = low_income, y = gdp_growth, color = as.factor(year))) +
  geom_point() +
  theme_minimal() +
  labs(x = "Low-income countries", y = "gdp_growth", color = "Years")+
  scale_x_continuous(breaks = c(0,1))
```
In @fig-gdp-growth, I have shown the gdp growth rate for the low-income and the high-income countries. On the x-axis, when x is equal to 0, it means that the county is a high-income country and it equals to 1 when the country is a low-income country. From @fig-gdp-growth, we could see that in both 2020 and 2021, each year the high income countries have similar gdp growth rates. Which in low-income countries the gdp growth rates more varies across each country. Lastly in both low and high income situations, the overall gdp growth in 2021 are higher than in 2020.

\newpage

```{r}
#| label: fig-ur
#| fig-cap: Relationship between differenct level of income countries versus labour force participation rate
#| echo: false
#| warning: false
#| message: false
data_na |>
  ggplot(mapping = aes(x = low_income, y = unem_rate, color = as.factor(year))) +
  geom_point() +
  theme_minimal() +
  labs(x = "Low-income countries", y = "Unemployment Rate", color = "Years")+
  scale_x_continuous(breaks = c(0,1))
```

From @fig-ur, the low-income countries seems to have lower unemployment rate than in high-income countries. Also, in both 2020 and 2021, the unemployment rates for both low and high income countries have similar results.  

Lastly, the data presented in @fig-inflation, @fig-gdp-growth, and @fig-ur, I observed that high-income countries shared similar economic performance trends. Although not necessarily always better in all economic indicators than the low-income countries, however, these countries generally maintain a more stable economic condition, with only a few exceptions or outliers noted. Therefore, I will be expecting the data to be normally distributed. Also the overall economic performance in 2021 are better than in 2020. 

A more detailed relationship between each variable could be found in @fig-relationship.


\newpage


# Model

The model is set up by using R [@citeR] with the help of the following packages, `tidyverse` package of @tidyverse, `readr` package of @readr, `dplyr` package of @dplyr, `dataverse` package of @dataverse, `stargazer` and package of  @stargazer, and `broom` package of @broom.
Given that countries are classified as low or high income based on their GNI per capita, `gni_per_cap`, I will use this as the dependent variable in a multiple linear regression model to analyze the impact of various factors on `gni_per_cap` across these groups. I will explore how different variables affect `gni_per_cap` in both low and high-income countries. Specifically, I expect that low-income countries will experience larger negative effect from unemployment and inflation rates, as well as experience negative or slower GDP growth. This analysis aims to provide insights into the economic challenges faced by these countries and inform potential policy impacts. 

\newpage

```{r}
#| echo: false
#| warning: false
#| message: false

lowincome <- data_na %>% filter(low_income == 1)
highincome <- data_na %>% filter(low_income == 0)
lowmodel <- glm(gni_per_cap ~ gdp_growth + population + unem_rate + lf_par_rate + trade_per + inflation, 
             family = gaussian(link = "identity"), data = lowincome)

stargazer(lowmodel, type = "text",
          title = "Summary of Low Income Model",
          align = TRUE,
          out = "model_summary.txt")

```
From the Summary of Low Income Model above, we could see the following situations: 

The GDP growth (`gdp_growth`) coefficient is -25.093 with a standard error of 21.917. This suggests there is a negative relationship between GDP growth and GNI per capita, though it is not statistically significant as its high p-value. Then the Unemployment rate (`unem_rate`) coefficient is -23.374 with a standard error of 34.236. This indicates a negative relationship between the unemployment rate and GNI per capita, which is also not statistically significant. Lastly the Inflation (`inflation`) coefficient is -7.082 with a standard error of 8.800, indicating a negative relationship between inflation and GNI per capita, though this relationship is not statistically significant.

The Labor force participation rate (`lf_par_rate`) coefficient is -52.818 with a standard error of 13.104, and it is significant at the p<0.01 level. This significant negative relationship shows that higher labor force participation is connected with lower GNI per capita among the low-income countries in this analysis, which might suggest underemployment or low-quality jobs in these economies.
The Trade percentage of GDP (`trade_per`) coefficient is 11.697 with a standard error of 3.992, significant at the p<0.01 level. This significant positive relationship suggests that a higher trade percentage of GDP is associated with higher GNI per capita, indicating the beneficial impact of trade on national income in low-income countries.

The Population (`population`) coefficient is different than the other variables. As its coefficient is extremely small and not statistically significant, suggesting that population size, has no impact on GNI per capita in this model.



```{r}
#| echo: false
#| warning: false
#| message: false
highmodel <- glm(gni_per_cap ~ gdp_growth + population + unem_rate + lf_par_rate + trade_per + inflation, 
             family = gaussian(link = "identity"), data = highincome)


stargazer(highmodel, type = "text",
          title = "Summary of High Income Model",
          align = TRUE,
          out = "model_summary.txt")
```

The high income-countries have an overall better performance than the low-income countries: 
The GDP Growth (`gdp_growth`) coefficient is 567.121 with a standard error of 244.367, significant at the p<0.05 level. This suggests a positive and statistically significant relationship between GDP growth and GNI per capita in high-income countries, indicating that economic growth positively impacts income levels. Then the Unemployment Rate (`unem_rate`) coefficient is -849.519 with a standard error of 350.854, significant at the p<0.05 level. This indicates a significant negative impact of higher unemployment rates on GNI per capita, suggesting that unemployment is a critical economic issue affecting income in high-income countries. The Labor Force Participation Rate (`lf_par_rate`) coefficient is 433.735 with a standard error of 229.487, significant at the p<0.1 level. This shows a positive relationship between labor force participation and GNI per capita, though the relationship is less statistically significant. The Trade Percentage of GDP (`trade_per`) coefficient is 59.875 with a standard error of 22.554, significant at the p<0.01 level. This shows a robust positive relationship, suggesting that trading contributes significantly to GNI per capita in high-income countries. Lastly, the Inflation (`inflation`) coefficient is -2,799.239 with a standard error of 600.123, significant at the p<0.01 level. This suggests a strong negative impact of inflation on GNI per capita, indicating that higher inflation rates are fluctuating income levels.

Same with the low-income situation, the Population (`population`) coefficient is extremely small and statistically insignificant, implying no clear impact of population size on GNI per capita within this model.


## Model set-up

To consider the effect of the 6 economic indicators on the level on low and high income countries' GNI per capita, I used the multiple linear regression to approach my data. The formula is as follows: 

\begin{align}
\text{gni\_per\_cap} &= \alpha + \beta_1 \times \text{gdp\_growth}_i + \beta_2 \times \text{population}_i + \beta_3 \times \text{unem\_rate}_i + \beta_4 \times \text{lf\_par\_rate}_i \\
&\quad + \beta_5 \times \text{trade\_per}_i + \beta_6 \times \text{inflation}_i
\end{align}

## Model Adjustment
For a better performance and interpretation of the model, I have applied the log transformation to my model. The log transformation is to reduce the skewness in variables that have a skewed distribution, such as `gni_per_cpa`, `gdp_growth`, `population`, etc. By transforming these variables, the distribution is more symmetric, which is closer to the normal distribution. The Log transformations also help stabilize the variance across different levels of input variables, reduces homoscedasticity.
The log transformation variables are the follows:
Log of `gni_per_cap` The dependent variable's log transformation can help in normalizing the distribution and making relationships more linear. 
The log of `gdp_growth` and `population` shared similar reasons. As economic growth rates and population can vary widely, and the impact on GNI per capita might scale logarithmically other than linearly.
The following formula is the updated log transformation on certain variables. 

\begin{align}
\log(\text{gni\_per\_cap}) &= \alpha + \beta_1 \times \log(\text{gdp\_growth}_i+1) + \beta_2 \times \log(\text{population}_i +1) \\ 
&\quad + \beta_3 \times \text{unem\_rate}_i + \beta_4 \times \text{lf\_par\_rate}_i \\
&\quad + \beta_5 \times \text{trade\_per}_i + \beta_6 \times \text{inflation}_i
\end{align}

```{r}
#| echo: false
#| warning: false
#| message: false
log_lowmodel <- glm(log(gni_per_cap) ~ log(gdp_growth + 1) + log(population + 1) + unem_rate + lf_par_rate + trade_per + inflation, family = gaussian(link = "identity"), data = lowincome)

stargazer(log_lowmodel, type = "text",
          title = "Summary of Low Income Model",
          align = TRUE,
          out = "model_summary.txt")
```

\newpage
```{r}
#| echo: false
#| warning: false
#| message: false
log_highmodel <- glm(log(gni_per_cap) ~ log(gdp_growth + 1) + log(population + 1) + unem_rate + lf_par_rate + trade_per + inflation,family = gaussian(link = "identity"), data = highincome)

stargazer(log_highmodel, type = "text",
          title = "Summary of High Income Model",
          align = TRUE,
          out = "model_summary.txt")
```


# Results

From the model we could see that in low-income countries the labour force participation rate has a significant negative value. Which in the high-income countries, the labour force participation rate has a significant positive value. This contrast may indicate that in low-income countries, increases in labor force participation may include low-quality, low-income jobs. Which are more unsecured during the COVID-19 crisis. Whereas in high-income countries, higher participation could be linked to more high-skilled roles. [@job]
The Trade as Percentage of GDP has positive impact and significant in both low and high income countries. Although the trade decreases sharply in 2020, however, the recovery in 2021 was more significant than the dropped in 2020. Causing the total trade flows are above the pre-pandemic levels.[@Trade]

The GDP Growth played an important role in the high-income countries and it has a significant and positive relationship. Possibly reflecting more developed financial systems and infrastructure which cause more growth into GNI per capita. However, it has insignificant negative relationship in the low-income countries. 

The unemployment rate and the inflation rate both have a significant and negative impact in the high-income countries. 
As the unemployment rate negatively impacts GNI per capita in high-income countries, where unemployment may lead to significant economic losses. The inflation is an important factor in high-income countries, however, due to more developed financial markets and economies, there are more ways to reduce the impact of inflation. Which in the low-income economies, the unemployment rate and the inflation rate have insignificant negative impact on GNI per capita. 

# Discussion

The results showed that the unemployment and inflation have significant negative impacts in high-income countries, as indicated by the strong and significant coefficients. These outcomes were unexpected, as the expectation was that these factors would have a larger negative effect in low-income countries than the high-income countries. It may be that high-income countries have a more developed financial markets and dependencies on stable economic indicators which are more sensitive to the fluctuations.

The negative coefficient for GDP growth in low-income countries was expected however, it was not significant. For high-income countries, GDP growth shows a positive and significant impact, indicating strong economic dynamics and growth directly improving national income levels. Also the high income countries showed a significant stronger positive relationship on labour force participation rate and trade as a share of GDP. 
From @fig-qq-low we could see that the observations were mostly separated and with multiple outliers which indicates that we have a low sample size also deviates from normality. Which shows that for the low-income countries model, it has a probability of leading to skewed or biased estimates.
From @fig-qq-high the points on the Q-Q plot fall approximately along a straight line, suggesting that the residuals follow a normal distribution. Which it also has less outliers compared with the low-income model. It also further confirm that the residuals are normally distributed.

## Weaknesses and next steps

There are several weaknesses of this study, which in the low-income countries model several important variables like GDP growth, unemployment, and inflation did not show statistically significant effects. This might be due to the lack of sample sizes and the high variations in the data. The second weakness was the lack of variables, the model did not include some relevant variables that affect GNI per capita, such as political stability, technology, and education levels, etc. Which both results might create biases onto the results.
The next step of the study might be on increase the data size, such as include more years or some micro-level data. 

# Additional data details
```{r}
#| label: fig-relationship 
#| fig-cap: Scatterplot matrix of each variables
#| echo: false
#| warning: false
#| message: false

pairs(~inflation + gdp_growth + population + unem_rate + lf_par_rate + trade_per + gni_per_cap, data = data_na)
```

```{r}
#| label: fig-qq-low 
#| fig-cap: Q-Q Plot of the low-income model
#| echo: false
#| warning: false
#| message: false

# Q-Q Plot
qqnorm(residuals(log_lowmodel), main = "Q-Q Plot of the low-income model")

```

```{r}
#| label: fig-qq-high
#| fig-cap: Q-Q Plot of the high-income model
#| echo: false
#| warning: false
#| message: false

# Q-Q Plot
qqnorm(residuals(log_highmodel), main = "Q-Q Plot of the high-income model")
```

\newpage


# References


